---
title: "Statistical Inference Final Assignment"
author: "Lou Solot"
date: "January 27, 2016"
output: html_document
---

## Overview

This report will consist of two parts. Part 1 will use simulation to compare the Central Limit Theorem with the exponential distribution. The means, standard deviations of the simulated samples will be compared with the theoretical values. 
Part 2 will perform tests using confidence intervals and hypothesis tests on tooth growth data

### Part 1
In this part I will take 1000 simulations of 40 exponentials. Lambda will be assumed to by 0.2 for the experiment. 
The exponential distribution mean is 1/lambda = `r 1/0.2` which is also the standard deviation
```{r, echo=TRUE, cache=FALSE, results='markup'}
means  <- NULL
nosim  <- 1000
n      <- 40
lambda <- 0.2

for (i in 1:nosim) {
        means <- c(means, mean(rexp(n, lambda)))
}
hist(rexp(1000,0.2), breaks = 30, main = "Histogram of 1000 Random Exponentials",xlab = "1000 Random Exponentials with lambda of 0.2",col = "darkblue")
```

<b>As you can see, the distribution of 1000 random exponentials is not gaussian</b>
<br></br>

```{r, echo=TRUE,cache=FALSE,results='markup'}
hist(means, breaks = 30, main = "Histogram of Exponential Means",col = "darkblue")
```

* The sample mean is mean(means) = `r mean(means)` 
* The theoretical mean is 1/lambda = `r 1/lambda`
* The distribution of the sample means is gaussian as the histogram above shows and is centered around the theoretical mean of 5 as predicted by the CLT

```{r,echo=TRUE,cache=FALSE,results='markup'}
vars <- NULL
nosim <- 1000
lambda <- 0.2
n <- 40
for (i in 1:nosim) {
        vars <- c(vars, var(rexp(n, lambda)))
}
hist(vars, breaks = 30, main = "Histogram of Exponential Variances", xlab = "Variances", col = "darkblue")
```


* The theoretical variance of the random distribution is 1/lambda^2 = `r 1/lambda^2`
* As shown in the figure above, the variance is (fairly) normally distributed, centered around the theoretical variance of 25
<br/>


### Part 2
```{r,echo=TRUE,results='markup',cache=FALSE}
library(datasets)
library(ggplot2)
data(ToothGrowth)
```

<br/><br/>

#### Let's explore the data


<b>Visualization of the tooth growth by dose, conditioned by supp</b></br>

```{r,echo=TRUE,results='markup',cache=FALSE}
options(warn = -1)
qplot(dose, len, data = ToothGrowth, facets = . ~ supp, geom = c("point", "smooth"), main = "Guinea Pig Tooth Growth by Vitamin C Dosage")
options(warn = 1)
```

<b>ToothGrowth data in long form</b><br/>

```{r,echo=TRUE,results='markup',cache=FALSE}
print(ToothGrowth)
```

<b>ToothGrowth data in wide form</b><br/>

```{r,echo=TRUE,results='markup',cache=FALSE}
## Had to add an id column because I couldn't get reshape to do what I needed otherwise
tlong <-ToothGrowth
tlong$id <- 1:10
twide <- reshape(timevar = "supp", direction = "wide", data = tlong, idvar = c("id","dose"))
twide$id <- NULL
print(twide)
print(mean(twide$len.VC))
print(mean(twide$len.OJ))
```

### Confidence Intervals
<br/><b>Let's run t-test confidence intervals since the samples are small. We will assume the null hypothesis that tooth growth will be the same with the vitamin C is delivered via Orange Juice or Ascorbic Acid</b>
<br/><br/>
1. Confidence interval for the 0.5 dosage
```{r, echo=TRUE, results='markup', cache=FALSE}
S1 <- twide[which(twide$dose == 0.5), "len.OJ"]
S2 <- twide[which(twide$dose == 0.5), "len.VC"]
t.test(S1, S2, var.equal = FALSE, conf.level = 0.95)
```

* This shows that we can reject the null hypothesis and conclude that Orange Juice is correlated with higher tooth growth than Ascorbic Acid

2. Confidence interval for the 1.0 dosage
```{r, echo=TRUE, results='markup', cache=FALSE}
S1 <- twide[which(twide$dose == 1.0), "len.OJ"]
S2 <- twide[which(twide$dose == 1.0), "len.VC"]
t.test(S1, S2, var.equal = FALSE, conf.level = 0.95)
```

* This shows that we can reject the null hypothesis and conclude that Orange Juice is correlated with higher tooth growth than Ascorbic Acid

2. Confidence interval for the 2.0 dosage
```{r, echo=TRUE, results='markup', cache=FALSE}
S1 <- twide[which(twide$dose == 2.0), "len.OJ"]
S2 <- twide[which(twide$dose == 2.0), "len.VC"]
t.test(S1, S2, var.equal = FALSE, conf.level = 0.95)
```

* This shows that we cannot reject the null hypothesis and conclude that Orange Juice and Ascorbic Acid show no meaningful statistical difference in tooth growth
